---
title: "BMI, Diet & Cancer Data Exploration"
author: "Gabriel Forn Cuní"
date: "25th September 2015"
output: 
  html_document:
    theme: cosmo
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages and defaults that well be used.

library(ggplot2)
library(dplyr)
library(reshape2)
library(grid)
library(gridExtra)
library(countrycode)
library(RColorBrewer)
library(rworldmap)

# Knittr defaults
knitr::opts_chunk$set(cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE)

# Global defaults
options(stringsAsFactors = FALSE)

# Plot defaults
theme_set(theme_minimal())

# rworldmap frame
par(mai=c(0,0,0.2,0),xaxs="i",yaxs="i")
```

# 0. Introduction

Obesity is a major risk factor for a number of chronic diseases, including 
diabetes mellitus, cardiovascular diseases, and cancer, and according to the 
World Health Organization, it has become a health challenge of epidemic 
proportions in the last few decades in the western and westernized societies.

Although highly criticized, the principal method for determining obesity in 
adults is through the Body Mass Index (BMI). BMI classification is as follows 
(from the [CDC](http://www.cdc.gov/obesity/adult/defining.html)):

Category |   Underweight  |    Healthy   |  Overweight  |    Obese
-------- | -------------- | ------------ | ------------ | ------------
  BMI    | Less than 18.5 | 18.5 to 24.9 | 25.0 to 29.9 | More than 30

In this project, I will analyze the relationship between nutritional behaviours 
(BMI & blood cholesterol levels) on worldwide populations and certain types of 
Cancer using information available in the Gapminder website 
(see <http://www.gapminder.org/data/>).

# 1. Worldwide BMI Exploration

I'll start exploring how the BMI dataset is structured and distributed.

```{r Load_the_data}
# Read the original data files
maleBMI <- read.csv("Indicator_BMI male ASM.csv", sep = ';')
femaleBMI <- read.csv('Indicator_BMI female ASM.csv', sep = ';')
```

## 1.1. Dataset structure

```{r BMI_structure}
str(maleBMI)
```

My dataset is structured in wide format. It contains data for 199 countries. 
The first column contains the name of the country and the rest values for the 
estimated average BMI in the country from the years 1980 to 2008. Male and 
female datasets are separated.

## 1.2. Dataset distribution

Let's take a grasp on the data and start analyzing the how the BMI is dispersed 
worldwide at the most recent time point available (2008), for males and females.

```{r echo = TRUE, BMI_summary}
summary(maleBMI$X2008)
summary(femaleBMI$X2008)
```

It seems that females have a slightly higher BMI than males. Let's plot the 
values as histograms.

```{r BMI_histograms}
plotMaleBMI <- ggplot(aes(x = X2008), data = maleBMI) +
  geom_histogram(binwidth = 0.25, color = '#ffffff', fill = '#333333') +
  xlab('Male BMI in 2008') +
  ylab('Number of Countries') +
  scale_x_continuous(limits = c(18,36), breaks = seq(18,36,1))
plotFemaleBMI <- ggplot(aes(x = X2008), data = femaleBMI) +
  geom_histogram(binwidth = 0.25, color = '#ffffff', fill = '#333333') +
  xlab('Female BMI in 2008') +
  ylab('Number of Countries') +
  scale_x_continuous(limits = c(18,36), breaks = seq(18,36,1))
grid.arrange(plotMaleBMI, plotFemaleBMI, ncol = 1)
```

The data distribution is very roughly normal, but the discreteness and the low 
number of points alter this pattern. I can see how there are a few more BMI 
values above 28 in the females plot. This information will be more evident using
boxplots.

```{r BMI_boxplots}
ggplot(aes(y = X2008, x = 'Males'), data = maleBMI) +
  geom_boxplot() + 
  geom_boxplot(aes(y = X2008, x = 'Females'), data = femaleBMI) +
  xlab('Worlwide BMI in 2008') +
  ylab('Number of Countries')
```

The boxplots confirm visually that generally **women have a slighter higher BMI 
than men** in this dataset. I wonder if this disparity has always existed. 
I can study how the BMI has evolved through the years 1980 to 2008. 

However, to do so I'll have to convert the data from wide format to long format.

## 1.3. BMI data reshape

```{r echo = TRUE, BMI_reshape}
# Add the gender column that we'll use later
maleBMI$Gender <- 'M'
femaleBMI$Gender <- 'F'

# Convert the data from wide to long format
maleBMI_long <- melt(maleBMI,
                     id.vars = c('Country', 'Gender'),
                     variable.name = 'Year',
                     value.name = 'BMI')
femaleBMI_long <- melt(femaleBMI,
                     id.vars = c('Country', 'Gender'),
                     variable.name = 'Year',
                     value.name = 'BMI')

# And merge male and female data.
healthData <- rbind(maleBMI_long, femaleBMI_long)

# Little quality of life improvements
healthData$Gender <- factor(healthData$Gender)
healthData$Year <- factor(healthData$Year, 
                          labels = c(1980:2008))
```

```{r BMI_housekeeping}
# Housekeeping, don't mind me
rm(list = ls(pattern = "^male"))
rm(list = ls(pattern = "^female"))
rm(list = ls(pattern = "^plot"))
```

Let's see how the dataset structure is now.

```{r BMI_long_structure}
str(healthData)
```

The new dataset contains 11542 observations (199 Countries, 2 Genders, 29 time 
points) of 3 variables: Country, Gender and Year of the BMI observation.

## 1.4. BMI evolution in the last decades

I can now use a scatterplot to see how the BMI has evolved worldwide since the 
80s to the 2008.

```{r BMI_evolution}
ggplot(aes(x = Year, y = BMI), data = healthData) +
  geom_point() +
  xlab('Year') +
  ylab('BMI in world countries') +
  scale_x_discrete(breaks = seq(1980,2008,2))
```

It seems that worlwide BMI it's increasing. Is men and women BMI increasing at 
the same ratio?

```{r BMI_gender_evolution}
ggplot(aes(x = Year, y = BMI), data = healthData) +
  geom_jitter(aes(pch = Gender), alpha = 1/10, color = '#cccccc') +
  geom_smooth(aes(group = Gender, color = Gender)) +
  scale_x_discrete(breaks = seq(1980,2008,2))
```

From this data, we can conclude that **from 1980 onwards the female BMI has 
been higher than the male BMI and both are increasing at a similar ratio 
worldwide.**

## 1.5. Categorical BMI classification

This data is *worrying*, because a lot of observations are above the BMI 
treshold for overweight, and a few points are above obesity treshold. This means
that the majority of population from some countries is obese. I'll add a 
category to the data to make obvious this fact.

```{r echo = TRUE, BMI_Classification}
healthData$BMIClassification <- cut(healthData$BMI, 
  breaks = c(0,18.5,25,30,50),
  labels = c('Underweight', 'Healthy', 'Overweight', 'Obese'))
```

And plot the data using different colors for this new classification.

```{r BMI_gender_evolution_limits}
ggplot(aes(x = Year, y = BMI, color = BMIClassification, pch = Gender), 
       data = healthData) +
  geom_jitter(alpha = 1) +
  scale_x_discrete(breaks = seq(1980,2008,2)) +
  scale_colour_brewer(palette='Reds')
```

I wonder which countries are above the obese treshold for the majority of its 
population.

```{r BMI_obese_countries}
unique(subset(healthData, BMI >= 30 & Year == 2008)$Country)
```

The results are very interesting, because they are counter-intuitive if you are 
not familiarized with the data. However, **these results relate perfectly to the
current obesity problem in the Pacific countries** 
(<https://en.wikipedia.org/wiki/Obesity_in_the_Pacific>).

This data would benefit a lot from being plotted on a world map. 

```{r BMI_world_map}
worldMap <- map_data(map = "world")
ggplot() +
  # base layer
  geom_map(data = worldMap, 
           map = worldMap, 
           aes(map_id = region, x = long, y = lat), 
           fill = "white", 
           colour = "black", 
           size = 0.25) +
  # Data layer
  geom_map(data = healthData, 
           map = worldMap, 
           aes(map_id = Country, fill = BMIClassification), 
           color = "white", 
           size=0.25)
```

Some countries are not displaying correctly, because the name of some countries 
do not match to the map names. Let's see who they are.

```{r echo = TRUE, country_name_fix}
# Determine which names are not matched to the map regions. 
# From the data to the map.
unique(healthData$Country)[!unique(healthData$Country) %in% 
                             unique(worldMap$region)]

# And form the map to the data.
unique(worldMap$region)[!unique(worldMap$region) %in% 
                          unique(healthData$Country)]

```

So the problem seems to be that the names on the geom_map regions are obsolete 
('USSR', 'Czechoslovakia'...), 'Republic' is abbreviate to 'Rep.' in our 
dataset, and other small errors. After some research, I found that the rworldmap
package will be more accurate and easier to use.

I will also do one more step to get the country names to ISO3 Country Codes in 
order to minimize mismatches, using the countrycode package.

```{r echo = TRUE, BMI_rworldmap}
# Create the country codes
healthData$CountryCode <- factor(countrycode(healthData$Country,
                                             'country.name', 
                                             'iso3c'))

# Join the data with the rworldmap map
mapBMI <- joinCountryData2Map(healthData,
                              joinCode = 'ISO3',
                              nameJoinColumn = 'CountryCode')
```

Apparently, one country code (totalling 58 observations) failed to match with a 
region in the map. In any case, that's much better than what we had before.

```{r BMI_rworldmap_plot}
# Plot the data
orangePalette <- brewer.pal(4,'Oranges')
mapCountryData(mapBMI, 
               nameColumnToPlot = 'BMIClassification',
               colourPalette = orangePalette,
               catMethod = 'categorical')
```

That's better. I want to zoom to the Pacific area to visualize the countries 
with high obesity rankings. 

```{r pacific_BMI_map_1}
# And plot the map within the limits.
mapCountryData(mapBMI, 
               nameColumnToPlot = 'BMIClassification',
               colourPalette = orangePalette,
               catMethod = 'categorical',
               xlim = c(-180,-140),
               ylim = c(-30,30))
```

Unfortunately, the countries are too small to see them clearly.

```{r pacific_BMI_map_2}
mapCountryData(mapBMI, 
               nameColumnToPlot = 'BMIClassification',
               colourPalette = orangePalette,
               catMethod = 'categorical',
               xlim = c(-170,-150),
               ylim = c(-20,10))
```


Two questions arise at this stage: 

1. Are those changes in BMI a consequence of bad dieting?

and 

2. How is this higher BMI correlated with cancer?

# 2. Is this increase in global BMI a consequence of the diet?

In short, **is this BMI increase an effect of fast food and bad diets?** 
Fast food is shown to contribute to high blood cholesterol levels. Let's analyze
how cholesterol levels are evolving worldwide.

## 2.1. Cholesterol Exploration

```{r Load_cholesterol_data}
maleCholesterol <- read.csv('Indicator_TC male ASM.csv', sep = ';')
femaleCholesterol <- read.csv('Indicator_TC female ASM.csv', sep = ';')
```

Exactly as did with the BMI, which distribution have the cholesterol levels 
worldwide?

```{r cholesterol_summary}
summary(maleCholesterol$X2008)
summary(femaleCholesterol$X2008)
```

Again, it seems that females have higher choelsterol than males.

```{r cholesterol_histograms}
plotMaleCholesterol <- ggplot(aes(x = X2008), data = maleCholesterol) +
  geom_histogram(binwidth = 0.1, color = '#ffffff', fill = '#333333') +
  xlab('Male Cholesterol blood levels in 2008 (mmol/L)') +
  ylab('Number of Countries') +
  scale_x_continuous(limits = c(3.5,5.75), breaks = seq(3.5,5.75,0.25))
plotFemaleCholesterol <- ggplot(aes(x = X2008), data = femaleCholesterol) +
  geom_histogram(binwidth = 0.1, color = '#ffffff', fill = '#333333') +
  xlab('Female Cholesterol blood levels in 2008 (mmol/L)') +
  ylab('Number of Countries') +
  scale_x_continuous(limits = c(3.5,5.75), breaks = seq(3.5,5.75,0.25))
grid.arrange(plotMaleCholesterol, plotFemaleCholesterol, ncol = 1)
```

## 2.2. Cholesterol Data Reshape

In order to study the evolution of blood cholesterol levels and integrate this 
data with the BMI data, I have to convert it from wide to long format.

```{r echo = TRUE, cholesterol_reshape}
# Add the gender column that we'll use later
maleCholesterol$Gender <- 'M'
femaleCholesterol$Gender <- 'F'

# Convert the data to long format
maleCholesterol_long <- melt(maleCholesterol,
                     id.vars = c('Country', 'Gender'),
                     variable.name = 'Year',
                     value.name = 'Cholesterol')
femaleCholesterol_long <- melt(femaleCholesterol,
                     id.vars = c('Country', 'Gender'),
                     variable.name = 'Year',
                     value.name = 'Cholesterol')

# Group Male and Female Data
cholesterolData <- rbind(maleCholesterol_long, femaleCholesterol_long)
cholesterolData$Gender <- factor(healthData$Gender)
cholesterolData$Year <- factor(healthData$Year, 
                          labels = c(1980:2008))

# And finally, integrate!
healthData <- merge(healthData, cholesterolData)
```

```{r cholesterol_housekeeping}
# Housekeeping, don't mind me
rm(list = ls(pattern = "^male"))
rm(list = ls(pattern = "^female"))
rm(list = ls(pattern = "^plot"))
rm(cholesterolData, worldMap, mapBMI, orangePalette)
```

## 2.3. Cholesterol evolution in the last decades

Is the cholesterol in blood concentration behaving equally as the BMI?

```{r cholesterol_evolution}
ggplot(aes(x = Year, y = Cholesterol), data = healthData) +
  geom_jitter(alpha = 1/10, color = '#cccccc') +
  geom_smooth(aes(group = Gender, color = Gender)) +
  scale_x_discrete(breaks = seq(1980,2008,2))
```

Interesting! **Although the worldwide BMI is increasing, the cholesterol in 
blood levels are decreasing!** I wasn't expecting this *at all*. Nevertheless, 
I'll continue with the analysis of the correlation between both parameters.

## 2.4. BMI & Cholesterol Analysis

Let's view it from a BMI classification point of view.

```{r cholesterol_BMI_evolution}
ggplot(aes(x = Year, y = Cholesterol), data = healthData) +
  geom_jitter(alpha = 1/10, color = '#cccccc') +
  scale_x_discrete(breaks = seq(1980,2008,2)) +
  geom_line(aes(group = interaction(BMIClassification,Gender), 
                color = BMIClassification, linetype = Gender), 
            stat = 'summary', 
            fun.y = median, 
            size = 0.75) +
  xlab('Year') +
  ylab('Cholesterol blood levels (mmol/L)')
```

Now, that makes a little more sense. **Cholesterol levels in blood are 
descending in healthy and overweight populations, but stay steady on Obese 
populations!**

Moreover, it seems that although the female cholesterol levels are higher 
worldwide, males have greater cholesterol levels in overweight and obese 
populations.

Taking into account these differences, is the BMI a good indicator for the 
cholesterol blood levels?

```{r BMI_cholesterol_boxplot}
ggplot(aes(x = BMIClassification, y = Cholesterol, fill = Gender), 
       data = subset(healthData, Year == 2008)) +
  geom_boxplot()
```

```{r echo = TRUE, BMI_cholesterol_correlation}
maleHealthData2008 <- subset(healthData, Year == 2008 & Gender == 'M')
femaleHealthData2008 <- subset(healthData, Year == 2008 & Gender == 'F')
cor(maleHealthData2008$BMI,maleHealthData2008$Cholesterol)
cor(femaleHealthData2008$BMI,femaleHealthData2008$Cholesterol)
cor(subset(healthData, Year == 2008)$BMI,
    subset(healthData, Year == 2008)$Cholesterol)
```

As expected, **there is a great correlation between BMI and cholesterol blood 
levels**, specially on males! So, although cholesterol blood levels worldwide 
are descending, they appear highly correlated with the BMI, which is increasing
worldwide.

At this stage, I realize that this data may be 'biased' because the nutritional
differences between highly-developed and non-developed countries, where 
malnutrition happens and, in consequence, lower cholesterol levels are expected.
In contrast, fast food is more typical of "westernized" countries. 

## 2.5. Country classification

To take into account the aforementioned differences in country development, I'll
use the data of GDP(PPP) per country. I decided to use this data because of the
following statement:

> The measure that most economists prefer, therefore, is GDP (PPP) ["GDP based 
> on purchasing power parity"] per capita. GDP (PPP) per capita compares 
> generalized differences in living standards on the whole between nations 
> because PPP takes into account the relative cost of living and the inflation
> rates of countries, rather than using just exchange rates, which may distort 
> the real differences in income. 
>
> [Source](http://bit.ly/1CJFb2F)

```{r Load_GDP_data}
CountryGDP <- read.csv('indicator gapminder gdp_per_capita_ppp.csv', sep = ';')
```

I decided to arbitrarilly define as "rich" countries the ones at the top 25% of
GDP(PPP), and as 'poor' countries the bottom 25% in relation to its GDP (PPP). 

Let's visualize the GDP distribution in a map for the 2008 data.

```{r results = 'hide', GDP_map_prep}
# Get the ISO country code.
CountryGDP$CountryCode <- factor(countrycode(CountryGDP$Country,
                                             'country.name', 
                                             'iso3c'))

# Join the data.
mapGDP <- joinCountryData2Map(CountryGDP,
                              joinCode = 'ISO3',
                              nameJoinColumn = 'CountryCode')

```

```{r GDP_map}
# Plot the map.
bluesPalette <- brewer.pal(7,'Blues')
mapCountryData(mapGDP, 
               nameColumnToPlot = 'X2008',
               colourPalette = bluesPalette)
```

The poorest countries are found in Africa.

I'll add the categoric classification for the GDP and integrate the data with 
the healthData from the year 2008.

```{r echo = TRUE, Reshape_and_merge_GDP_data}
# Get the 2008 data and rename the columns
GDP2008 <- CountryGDP[c('Country', 'X2008')]
names(GDP2008) <- c('Country', 'GDP')

# Add the categorical classification
GDP2008$GDPClassification <- cut(GDP2008$GDP, 
  breaks = c(min(GDP2008$GDP), 
            quantile(GDP2008$GDP, 0.25), 
            quantile(GDP2008$GDP, 0.75), 
            max(GDP2008$GDP)),
  labels = c('Poor', 'Middle', 'Rich'))

# There are not the same countries in the two datasets...
countriesWithBMI <- GDP2008$Country

# ... And I have to keep male and female BMI separated.
maleHealthData2008 <- subset(subset(healthData, 
                                    Gender == 'M' & Year == '2008'), 
                             Country %in% countriesWithBMI)
femaleHealthData2008 <- subset(subset(healthData, 
                                      Gender == 'F' & Year == '2008'), 
                               Country %in% countriesWithBMI)

maleHealthData2008 <- merge(GDP2008, maleHealthData2008, all=FALSE)
femaleHealthData2008 <- merge(GDP2008, femaleHealthData2008, all=FALSE)

# The final 2008 data frame
healthData2008 <- rbind(maleHealthData2008, femaleHealthData2008)
```

```{r GDP_housekeeping}
# Housekeeping, don't mind me.
rm(countriesWithBMI, maleHealthData2008, femaleHealthData2008, GDP2008,
   bluesPalette, mapGDP)
```

## 2.5.1 Country GDP & Cholesterol

Let's do a simple scatterplot of Cholesterol levels vs GDP (PPP) factoring the 
Gender in.

```{r Cholesterol_GDP_scatter}
ggplot(aes(x = GDP, y = Cholesterol, color = GDPClassification, pch = Gender), 
       data = healthData2008) +
  geom_point()
```

This seems to be an exponential relationship. Let's alter the scale.

```{r Cholesterol_GDP_scatter_scale}
ggplot(aes(x = GDP, y = Cholesterol, color = GDPClassification, pch = Gender), 
       data = healthData2008) +
  geom_point() +
  scale_x_log10()
```

Yeah! But this boxplots would suit more this type of graph! I'll also mark the 
line between Healthy and Borderline high cholesterol levels 
([Source](http://mayocl.in/1KxcjuH)).

```{r Cholesterol_GDP_boxplot_clean}
ggplot(aes(x = GDPClassification, y = Cholesterol), 
       data = subset(healthData2008, !is.na(GDPClassification))) +
  geom_boxplot(alpha = 1/5, aes(fill = Gender)) +
  #scale_x_log10() +
  #geom_smooth(aes(group = Gender, color = Gender)) +
  xlab('Country GDP(PPP)') +
  ylab('Average blood cholesterol levels (mmol/L)') +
  geom_hline(yintercept=(5.2), linetype = 2, color = '#333333', alpha = 1/2)
```

Let's quantify that relationship.

```{r Cholesterol_GDP_correlation}
cor(healthData2008$GDP, healthData2008$Cholesterol)
```

So we can conclude that **blood cholesterol levels are highly correlated with 
country development**. This fact may affect the relationship between cholesterol
and BMI levels that we found before.

## 2.5.2 Country GDP & BMI

I wonder if the estimated country BMI is equally associated with its country
development level.

```{r BMI_GDP_graphics}
ggplot(aes(x = GDP, y = BMI, pch = Gender, color = GDPClassification), 
       data = healthData2008) +
  geom_point()

ggplot(aes(x = GDP, y = BMI, color = GDPClassification, pch = Gender), 
       data = healthData2008) +
  geom_point() +
  scale_x_log10() +
  geom_smooth(aes(group = Gender)) +
  geom_hline(yintercept = c(25,30), alpha = 1/5, linetype = 2)

ggplot(aes(x = GDPClassification, y = BMI, color = Gender), 
       data = subset(healthData2008, !is.na(GDPClassification))) +
  geom_boxplot()
```

Although poor countries tend to have lower BMI, the difference between middle 
and rich countries is not clear.

# 2.5.3. Country GDP, BMI & Cholesterol levels

Now it's time to take into account the three parameters. In this context, it 
means that I will only consider the data from "developed" countries, and ignore
the one from the "poorest" countries because I consider it unreliable (for the
nutritional reasons explained above, and the medical that will explain below).

As I'm gonna be subsetting for this parameter in almost every analysis from here
onwards, I decided to create a simple function for accessing it more easily. 
Similarly, I'm gonna do the same extremely easy functions with male and female
subsetting.

```{r echo = TRUE, custom_GDP_function}

# Returns the dataframe without the Poor and NA classifications for the 
# GDPClassification parameter. 
nonPoor <- function(dataframe) {
  return(subset(dataframe, 
         GDPClassification == 'Middle' | GDPClassification == 'Rich'))
}

# Returns the dataframe without subsetted by males of the Gender parameter.
males <- function(dataframe) {
  return(subset(dataframe, Gender == 'M'))
}
# Returns the dataframe without subsetted by males of the Gender parameter.
females <- function(dataframe) {
  return(subset(dataframe, Gender == 'F'))
}
```


```{r three_variables_GDP}
ggplot(aes(x = BMIClassification, y = Cholesterol), 
       data = nonPoor(subset(healthData2008, Year == 2008))) +
  geom_boxplot(aes(fill = Gender), alpha = 1/5) +
  xlab('Country BMI Classification') +
  ylab('Average country blood cholesterol levels (mmol/L)') +
  ggtitle('Relationship between BMI classification and Cholesterol\n
          blood levels in "non-poor" world populations. 2008') +
  geom_hline(yintercept=(5.2), linetype = 2, color = '#333333', alpha = 1/2)
```

**When we ignore the data from the "poorest" countries, the correlation between 
BMI and cholesterol dissapears.** 

# 3. Cancer

Finally, let's integrate what we have with cancer incidence data. 
Let's start by taking a look at the cancer data and its distribution.

## 3.1. Cancer data structure

```{r Load_cancer_data}
Cancer2002 <- read.csv('Indicator_Cancer_All.csv', sep = ';')
```

```{r Cancer_data_structure}
str(Cancer2002)
```

The dataset contains the number of cancer cases for 100.000 habitants of 
different cancer types at several countries from the year 2002. I will use the 
2002 country data for the next analysis in consequence.

## 3.2. Cancer data distribution

```{r cancer_histograms}
ggplot(aes(x = value), data = melt(Cancer2002)) + 
    facet_wrap(~variable, scales = 'free_x') + 
    geom_histogram()

summary(Cancer2002)
```

The distribution for the majority of cancer incidences appears to be somewhat
long-tailed. Again, I worry that the development and, in consequence, medical
advancement of the countries affects the number of cancers detected. For this,
I'll take into account the country GDP in the next analyses too.

## 3.3. Cancer data integration

Next, I'll integrate the number of cancer cases with the health and economic 
2002 data. 

```{r echo = TRUE, cancer_data_reshape}
# This time I need the 2002 data. First the economic parameters.
GDP2002 <- CountryGDP[c('Country', 'X2002')]
names(GDP2002) <- c('Country', 'GDP')
GDP2002$GDPClassification <- cut(GDP2002$GDP, 
                                 breaks = c(min(GDP2002$GDP), 
                                            quantile(GDP2002$GDP, 0.25), 
                                            quantile(GDP2002$GDP, 0.75), 
                                            max(GDP2002$GDP)),
                                 labels = c('Poor', 'Middle', 'Rich'))

# And now the health data. I separate men and women because there are specific
# cancer types for the genres.
# First, subset the BMI & Cholesterol levels
maleHealthData2002 <- subset(males(healthData), Year == 2002)
# Add the economic data
maleHealthData2002 <- merge(maleHealthData2002, GDP2002, all = FALSE)
# Add the cancer incidence data
maleHealthData2002 <- merge(maleHealthData2002, 
                            Cancer2002[c('Country', 'Prostate.Male', 
                                         'Colon.Male', 'Liver.Male', 
                                         'Lung.Male', 'Stomach.Male')], 
                            all = FALSE)
# Change the column names so it matches with the female data
names(maleHealthData2002) <- c('Country', 'Gender', 'Year', 'BMI', 
   'BMIClassification', 'CountryCode', 'Cholesterol', 'GDP', 
   'GDPClassification', 'Prostate', 'Colon', 'Liver', 'Lung', 'Stomach')
# Add the Breast and Cervix with NA
maleHealthData2002$Cervix <- NA
maleHealthData2002$Breast <- NA

# The same for female data
femaleHealthData2002 <- subset(females(healthData), Year == 2002)
femaleHealthData2002 <- merge(femaleHealthData2002, GDP2002, all = FALSE)
femaleHealthData2002 <- merge(femaleHealthData2002, 
                              Cancer2002[c('Country', 'Breast.Cancer', 'Cervix',
                                           'Colon.Female', 'Liver.Female', 
                                           'Lung.Female', 'Stomach.Female')], 
                              all = FALSE)

names(femaleHealthData2002) <- c('Country', 'Gender', 'Year', 'BMI', 
   'BMIClassification', 'CountryCode', 'Cholesterol', 'GDP', 
   'GDPClassification', 'Breast', 'Cervix', 'Colon', 'Liver', 'Lung', 'Stomach')
femaleHealthData2002$Prostate <- NA

# And joint the two datasets
healthData2002 <- rbind(maleHealthData2002, femaleHealthData2002)

```

```{r cancer_housekeeping}
# Housekeeping, don't mind me.
rm(Cancer2002, GDP2002, maleHealthData2002, femaleHealthData2002)
```

## 3.4. Non-gastrointestinal cancers

There are 3 gender-specific, non-gastrointestinal cancers in the dataset: 
brest, cervix, and prostate cancer. I'll analyze them first because they should
be less correlated with nutriotional data than gastrointentestinal cancers.

## 3.4.1. Prostate Cancer

Prostate Cancer only affects males so it should be more easy to work with. As 
reasoned above, I'll take into account the country GDP because it is expected 
that more cancers are detected in richer countries because of better health 
systems.

```{r Prostate_exploration}
prostateBMI <- ggplot(aes(x = BMI, y = Prostate), 
                 data = males(healthData2002)) +
  geom_point() +
  facet_wrap(~GDPClassification, ncol = 1)

prostateCholesterol <- ggplot(aes(x = Cholesterol, y = Prostate), 
                 data = males(healthData2002)) +
  geom_point() +
  facet_wrap(~GDPClassification, ncol = 1)

grid.arrange(prostateBMI, prostateCholesterol, ncol = 2)
```

There actually seems to be certain **correlation between the BMI and cholesterol 
levels and the prostate cancer incidences**, which I didn't expect.

```{r Prostate_correlation}
prostateBMI <- ggplot(aes(x = BMI, y = Prostate), 
       data = nonPoor(males(healthData2002))) +
  geom_point() +
  geom_smooth(method = 'lm') +
  coord_cartesian(ylim = c(0,130))

prostateCholesterol <- ggplot(aes(x = Cholesterol, y = Prostate), 
       data = nonPoor(males(healthData2002))) +
  geom_point() +
  geom_smooth(method = 'lm') +
  coord_cartesian(ylim = c(0,130))

grid.arrange(prostateBMI, prostateCholesterol, ncol = 2)
rm(prostateBMI, prostateCholesterol)
```

A rapid search confirms this findings:

> Epidemiological studies have associated high blood-cholesterol levels with an
increased risk of Prostate Cancer. 
>
> [Source](http://bit.ly/1KxVL5J)

## 3.4.2. Breast and Cervix Cancers.

Breast and cervix cancers only affect women. Let's see its correlation with BMI
and cholesterol levels.

```{r Breast_cancer}

breastBMI <- ggplot(aes(x = BMI, y = Breast), 
                 data = nonPoor(females(healthData2002))) +
  geom_point() +
  facet_wrap(~GDPClassification, ncol = 1)

breastCholesterol <- ggplot(aes(x = Cholesterol, y = Breast), 
                  data = nonPoor(females(healthData2002))) +
  geom_point() +
  facet_wrap(~GDPClassification, ncol = 1)

grid.arrange(breastBMI, breastCholesterol, ncol = 2)
```

It appears that choelsterol levels affect Breast cancer incidence, but BMI does
not. 

```{r Breast_correlation}
breastBMI <- ggplot(aes(x = BMI, y = Breast), 
       data = nonPoor(females(healthData2002))) +
  geom_point() +
  geom_smooth(method = 'lm') +
  coord_cartesian(ylim = c(0,110))

breastCholesterol <- ggplot(aes(x = Cholesterol, y = Breast), 
       data = nonPoor(females(healthData2002))) +
  geom_point() +
  geom_smooth(method = 'lm') +
  coord_cartesian(ylim = c(0,110))

grid.arrange(breastBMI, breastCholesterol, ncol = 2)
rm(breastBMI, breastCholesterol)
```

We can see how **cholesterol is highly correlated with Breast Cancer 
incidence**. And again, it seems that we are supported by science.

> High cholesterol levels may increase a woman's risk of developing breast 
> cancer, a large new British study reports.
>
> [Source](http://wb.md/1LRnsnK)

```{r Cervix_cancer}
cervixBMI <- ggplot(aes(x = BMI, y = Cervix), 
                 data = nonPoor(females(healthData2002))) +
  geom_point() +
  facet_wrap(~GDPClassification, ncol = 1)

cervixCholesterol <- ggplot(aes(x = Cholesterol, y = Cervix), 
                 data = nonPoor(females(healthData2002))) +
  geom_point() +
  facet_wrap(~GDPClassification, ncol = 1)

grid.arrange(cervixBMI, cervixCholesterol, ncol = 2)
rm(cervixBMI,cervixCholesterol)
```

In contrast, cervix Cancer does not appear affected by this parameters.

## 3.4.3. Lung Cancer

The last non-gastrointestinal cancer that we have incidence data is Lung cancer.
Lung cancer affects both men and women, so we can study both groups in this
analysis.

```{r Lung_cancer}
allLungBMI <- ggplot(aes(x = BMI, y = Lung), 
                 data = nonPoor(healthData2002)) +
  geom_point() +
  ggtitle('BMI & Lung Cancer both genders\n')

lungBMI <- ggplot(aes(x = BMI, y = Lung), 
                 data = nonPoor(healthData2002)) +
  geom_point() +
  facet_wrap(~Gender, ncol = 2) +
  ggtitle('BMI & Lung Cancer per gender\n')

allLungCholesterol <- ggplot(aes(x = Cholesterol, y = Lung), 
                 data = nonPoor(healthData2002)) +
  geom_point() +
  ggtitle('Cholesterol & Lung Cancer both genders\n')

lungCholesterol <- ggplot(aes(x = Cholesterol, y = Lung), 
                 data = nonPoor(healthData2002)) +
  geom_point() +
  facet_wrap(~Gender, ncol = 2) +
  ggtitle('Cholesterol & Lung Cancer per gender\n')

grid.arrange(allLungBMI, allLungCholesterol, lungBMI, lungCholesterol, ncol = 2)
rm(allLungBMI, allLungCholesterol, lungBMI, lungCholesterol)
```

Again, **cholesterol levels are correlated with higher Lung Cancer incidence in
male and female populations**, but specially in men.

Let's model this correlation in "non-poor" countries.

```{r Lung_cholesterol_correlation}

ggplot(aes(x = Cholesterol, y = Lung, pch = Gender), 
       data = nonPoor(healthData2002)) +
  geom_point() +
  geom_smooth(aes(color = Gender), method = 'lm') +
  coord_cartesian(ylim = c(0,100))

```

Again, it appears that we are somewhat backed by science. In a Hawaii study, it
was found that:

> The results showed a significant positive association of dietary cholesterol 
> and the risk of lung cancer in men, but not in women.
>
> [Source](http://www.ncbi.nlm.nih.gov/pubmed/3195565)

## 3.5. Gastrointestinal cancers

It's now time to analyze the incidence of gastrointestinal-related cancers. It 
can be hypothesized that gastrointestinal-related cancers such as Colon, Stomach
or Liver cancers may have more relationship with dietary and nutritional 
parameters.

To speed up the process of analyzing those cancers, I'm gonna use a for loop to 
make and scatterplot of all of them and focus on the most insteresting ones.

```{r echo = TRUE, gastrointestinal_cancers}

gastrointestinalCancers <- c('Stomach', 'Colon', 'Liver')

for (cancer in gastrointestinalCancers) {
  # NOTE: I should do a for loop (parameter in parameters), but I don't get the
  # x axis to work, although the rest is all ok.
  bothBMI <- ggplot(aes(x = BMI, 
                        y = nonPoor(healthData2002)[[cancer]]), 
      data = nonPoor(healthData2002)) +
      geom_point() +
      xlab('BMI') +
      ylab(cancer)
  sexBMI <- bothBMI + facet_wrap(~Gender, ncol = 2)
  
  bothCholesterol <- ggplot(aes(x = Cholesterol,
                         y = nonPoor(healthData2002)[[cancer]]), 
      data = nonPoor(healthData2002)) +
      geom_point() +
      xlab('Cholesterol') +
      ylab(cancer)
  sexCholesterol <- bothCholesterol + facet_wrap(~Gender, ncol = 2)
  
  grid.arrange(bothBMI, bothCholesterol, sexBMI, sexCholesterol, ncol = 2)
}

```

```{r gastrointestinal_housekeeping}
# Housekeeping, don't mind me
rm(list = ls(pattern = "^both"))
rm(list = ls(pattern = "^sex"))
rm(gastrointestinalCancers, cancer)
```


There seems to be a trend in the Colon cancers with our nutritional parameters. 
I also notice some outliers in the liver cancer plots, but it doesn't appear to 
mask any trend.

Let's focus on the colon cancer correlation.

```{r colon_cancer_plots}
colonBMI <- ggplot(aes(x = BMI, y = Colon, color = Gender, pch = Gender), 
                   data = nonPoor(healthData2002)) +
  geom_point() +
  geom_smooth(method = 'lm', alpha = 1/10)

colonCholesterol <- ggplot(aes(x = Cholesterol, y = Colon, color = Gender, 
                               pch = Gender), data = nonPoor(healthData2002)) +
  geom_point() +
  geom_smooth(method = 'lm', alpha = 1/10)
grid.arrange(colonBMI, colonCholesterol, ncol = 2)
rm(colonBMI, colonCholesterol)
```

So **although BMI does not appear to have an effect in the number of 
gastrointestinal cancers, high blood cholesterol levels do**. 

## 3.6. Summary Correlation Cholesterol and Cancer Incidence

As a summary, let's plot the scatterplot in which we found a correlation with
cholesterol.

```{r fig.width = 10, cholesterol_cancers}

prostate <- ggplot(aes(x = Cholesterol, y = Prostate), 
       data = males(nonPoor(healthData2002))) +
  geom_smooth(method = 'lm', alpha = 1/8, color = '#000000') +
  geom_point(color = '#34495e', alpha = 1/4) +
  coord_cartesian(ylim = c(0,130)) + 
  geom_vline(xintercept = 5.2, linetype = 2, alpha = 1/3) +
  xlab('Population Cholesterol blood levels (mmol/L)') +
  ylab('Number of Prostate cancer cases per\n100,000 habitants') +
  ggtitle('Relationship between Cholesterol blood levels and Prostate Cancer
           cases in men world populations. 2002\n') +
  theme_minimal(base_size = 9)

breast <- ggplot(aes(x = Cholesterol, y = Breast), 
       data = females(nonPoor(healthData2002))) +
  geom_smooth(method = 'lm', alpha = 1/8, color = '#000000') +
  geom_point(color = '#34495e', alpha = 1/4) +
  coord_cartesian(ylim = c(0,110)) + 
  geom_vline(xintercept = 5.2, linetype = 2, alpha = 1/3) +
  xlab('Population Cholesterol blood levels (mmol/L)') +
  ylab('Number of Breast cancer cases per\n100,000 habitants') +
  ggtitle('Relationship between Cholesterol blood levels and Breast Cancer
           cases in women world populations. 2002\n') +
  theme_minimal(base_size = 10)

lung <- ggplot(aes(x = Cholesterol, y = Lung), data = nonPoor(healthData2002)) +
  geom_smooth(method = 'lm', alpha = 1/8, color = '#000000') +
  geom_smooth(aes(color = Gender), method = 'lm', alpha = 1/8) +
  geom_point(aes(pch = Gender), color = '#34495e', alpha = 1/4) +
  coord_cartesian(ylim = c(0,75)) + 
  geom_vline(xintercept = 5.2, linetype = 2, alpha = 1/3) +
  xlab('Population Cholesterol blood levels (mmol/L)') +
  ylab('Number of Lung cancer cases per\n100,000 habitants') +
  ggtitle('Relationship between Cholesterol blood levels and Lung Cancer
          cases in world populations. 2002\n') +
  theme_minimal(base_size = 10)

colon <- ggplot(aes(x = Cholesterol, y = Colon),
                data = nonPoor(healthData2002)) +
  geom_smooth(method = 'lm', alpha = 1/8, color = '#000000') +
  geom_smooth(aes(color = Gender), method = 'lm', alpha = 1/8) +
  geom_point(aes(pch = Gender), color = '#34495e', alpha = 1/4) +
  coord_cartesian(ylim = c(0,60)) + 
  geom_vline(xintercept = 5.2, linetype = 2, alpha = 1/3) +
  xlab('Population Cholesterol blood levels (mmol/L)') +
  ylab('Number of Colon cancer cases per\n100,000 habitants') +
  ggtitle('Relationship between Cholesterol blood levels and
          Colon Cancer cases world populations. 2002\n') +
  theme_minimal(base_size = 10)

grid.arrange(prostate, breast, lung, colon, ncol = 2)
rm(prostate, breast, lung, colon)

```

## 3.7. BMI, Cholesterol & General Cancer Incidence

Finally, let's take a step back and study the simple relationship between BMI 
and cholesterol and general cancer incidence. 

### 3.7.1. Data reshape

To do so, I'll convert the data from wide to long format, and add two new 
columns, in which I round the BMI and cholesterol levels to certain decimals 
to easily group the data and plot histograms.

```{r echo = TRUE, All_cancer_reshape}
# Remove the columns we don't want
healthData2002Long <- nonPoor(healthData2002)[-c(3,5,6,8,9)]
# Convert the data to long format
healthData2002Long <- melt(healthData2002Long,
                        id.vars = c('Country', 'Gender', 'BMI', 'Cholesterol'),
                        variable.name = 'TypeOfCancer',
                        value.name = 'NumberOfCases')
# Round BMI to the nearest 0.5
healthData2002Long$BMIRounded <- round(healthData2002Long$BMI*2,0)/2
# Round Cholesterol to the nearest 0.1
healthData2002Long$CholesterolRounded <- round(healthData2002Long$Cholesterol,1)
```

### 3.7.2. Cancer Incidence Distribution

Let's plot the data to view the distribution.

```{r All_cancer_histograms}
cancersBMI <- ggplot(aes(x = BMIRounded, y = NumberOfCases, 
                         fill = TypeOfCancer), data = healthData2002Long) +
  geom_bar(stat = 'identity')
cancersChol <- ggplot(aes(x = CholesterolRounded, y = NumberOfCases, 
                          fill = TypeOfCancer), data = healthData2002Long) +
  geom_bar(stat = 'identity')

grid.arrange(cancersBMI, cancersChol, ncol = 1)
rm(cancersBMI, cancersChol)
```

### 3.7.2. Cancer Incidence Normalization

That's nice, but we have to normalize the data, taking into account the number
of cases for each BMI and cholesterol level occurrence. I'll use dplyr to group 
the data. It is important that we ignore the NAs caused by gender-specficic 
cancers here.

```{r echo = TRUE, All_cancer_normalization}

BMICancerSummary <- healthData2002Long %>%
  group_by(BMIRounded, Gender, TypeOfCancer) %>%
  summarize(meanNumberOfCancers = mean(NumberOfCases, na.rm = TRUE),
            n = n()) %>%
  arrange(BMIRounded)

CholCancerSummary <- healthData2002Long %>%
  group_by(CholesterolRounded, Gender, TypeOfCancer) %>%
  summarize(meanNumberOfCancers = mean(NumberOfCases, na.rm = TRUE),
            n = n()) %>%
  arrange(CholesterolRounded)

# As an additional step, I'll change all the NaN values to 0.
BMICancerSummary$meanNumberOfCancers[
  is.nan(BMICancerSummary$meanNumberOfCancers)] <- 0
CholCancerSummary$meanNumberOfCancers[
  is.nan(CholCancerSummary$meanNumberOfCancers)] <- 0

```

Let's plot the normalized data.

```{r All_cancer_normalized}
plotBMICancerSummary <- ggplot(aes(x = BMIRounded, y = meanNumberOfCancers, 
                             fill = TypeOfCancer), 
                         data = BMICancerSummary) +
  geom_bar(stat = 'identity')
plotCholCancerSummary <- ggplot(aes(x = CholesterolRounded, 
                              y = meanNumberOfCancers, fill = TypeOfCancer), 
                          data = CholCancerSummary) +
  geom_bar(stat = 'identity')

grid.arrange(plotBMICancerSummary, plotCholCancerSummary, ncol = 1)
rm(plotBMICancerSummary, plotCholCancerSummary)
```

That makes more sense. It is interesting how the mean number of cases increases
with the cholesterol levels. However, here we have the data for male and female
population grouped, and there are specific cancers for each gender in the plot.
Let's view its gender in more detail. 

A part from the number of cancers per parameter, I'd like to see the composition
of this cancers, to corroborate that some types of cancers appear more at 
different BMI and Cholesterol, independently of the total number of cancer 
incidence.

### 3.7.3. Male Cancer Incidence

```{r Male_cancer_histograms}
maleBMICancerSummary <- ggplot(aes(x = BMIRounded, y = meanNumberOfCancers, 
                             fill = TypeOfCancer), 
                         data = males(BMICancerSummary)) +
  geom_bar(stat = 'identity')
maleCholCancerSummary <- ggplot(aes(x = CholesterolRounded, 
                              y = meanNumberOfCancers, fill = TypeOfCancer), 
                          data = males(CholCancerSummary)) +
  geom_bar(stat = 'identity')
fillmaleBMICancerSummary <- ggplot(aes(x = BMIRounded, y = meanNumberOfCancers, 
                             fill = TypeOfCancer), 
                         data = males(BMICancerSummary)) +
  geom_bar(stat = 'identity', position = 'fill') 
fillmaleCholCancerSummary <- ggplot(aes(x = CholesterolRounded, 
                              y = meanNumberOfCancers, fill = TypeOfCancer), 
                          data = males(CholCancerSummary)) +
  geom_bar(stat = 'identity', position = 'fill')
grid.arrange(maleBMICancerSummary, maleCholCancerSummary, 
             fillmaleBMICancerSummary, fillmaleCholCancerSummary, ncol = 2)

rm(maleBMICancerSummary, maleCholCancerSummary, fillmaleBMICancerSummary,
   fillmaleCholCancerSummary)
```

These graphs give an overall perspective on the cancer relationship with BMI
and cholesterol levels in male populations. We can clearly see how, similar to
what we saw with the combined populations, the number cancer incidence rises at
higher cholesterol levels, and how the percentage of its cancer type evolves 
during this evolution. For example, it is clear here that colon cancer incidence
is higher at higher cholesterol levels, while the percentage of prostate cancer
to the rest of cancer types is not dependent of the cholesterol levels. 

### 3.7.4. Female population

```{r Female_cancer_histograms}
femaleBMICancerSummary <- ggplot(aes(x = BMIRounded, y = meanNumberOfCancers, 
                             fill = TypeOfCancer), 
                         data = females(BMICancerSummary)) +
  geom_bar(stat = 'identity')
femaleCholCancerSummary <- ggplot(aes(x = CholesterolRounded, 
                              y = meanNumberOfCancers, fill = TypeOfCancer), 
                          data = females(CholCancerSummary)) +
  geom_bar(stat = 'identity')
fillfemaleBMICancerSummary <- ggplot(aes(x = BMIRounded, 
                                         y = meanNumberOfCancers, 
                             fill = TypeOfCancer), 
                         data = females(BMICancerSummary)) +
  geom_bar(stat = 'identity', position = 'fill') 
fillfemaleCholCancerSummary <- ggplot(aes(x = CholesterolRounded, 
                              y = meanNumberOfCancers, fill = TypeOfCancer), 
                          data = females(CholCancerSummary)) +
  geom_bar(stat = 'identity', position = 'fill')
grid.arrange(femaleBMICancerSummary, femaleCholCancerSummary, 
             fillfemaleBMICancerSummary, fillfemaleCholCancerSummary, ncol = 2)

rm(femaleBMICancerSummary, femaleCholCancerSummary, fillfemaleBMICancerSummary,
   fillfemaleCholCancerSummary)
```

The results for the female population are highly similar to the ones obtained 
for the male population. It is interesting to see how the breast cancer 
percentage in comparison to other cancer types rises at higher cholesterol 
levels, while the stomach cancer percentage diminishes.

### 3.8. Cholesterol and Common Cancers Types in Men and Women

To finish this data exploration study, I'd like to visualize the evolution of 
different cancer types common to both men and women with the blood cholesterol
levels, as it's the most striking relationship that I found in this study. To
do so, I'll group the data once again, this time without separating by gender.

```{r echo = TRUE, Cholesterol_Cancer_Summary}
AllCholCancerSummary <- subset(CholCancerSummary,  TypeOfCancer != 'Cervix' & 
                       TypeOfCancer != 'Breast' & TypeOfCancer != 'Prostate')
AllCholCancerSummary <-  AllCholCancerSummary %>%
  group_by(CholesterolRounded, TypeOfCancer) %>%
  summarize(meanNumberOfCancers = mean(meanNumberOfCancers),
            n = n()) %>%
  arrange(CholesterolRounded)
```

And plot the data.

```{r Common_Cancers_histogram}
ggplot(aes(x = CholesterolRounded, 
           y = meanNumberOfCancers, 
           fill = TypeOfCancer), 
       data = AllCholCancerSummary) +
  geom_bar(stat = 'identity') +
  scale_fill_brewer(palette = 'Pastel1') +
  geom_vline(xintercept = 5.2, linetype = 2, alpha = 1/3)
```

I think this plot summarizes perfectly the relationship that I was searching for
between nutritional parameters and cancer incidence. My hypothesis was that I 
would find this relationship with the BMI, as its the used indicator for 
obesity, but finally found that the cholesterol blood levels are a better
predictor for this data.

# 4. Final Plots and Summary

In this study I wanted to analyze if the increased obesity worldwide was related
to a worsened diet, and what effect would this have on the number cancer levels
at a population level. 

## Obesity is increasing worldwide.

```{r fig.width = 10, Final_Plot_1}

ggplot(aes(x = Year, y = BMI), data = healthData) +
  geom_jitter(alpha = 1/10, color = '#CCCCCC', aes(pch = Gender)) +
  geom_smooth(aes(group = Gender, color = Gender), alpha = 1/2) +
  scale_x_discrete(breaks = seq(1980,2008,2)) +
  geom_hline(yintercept = c(18.5,25,30), linetype = 2, alpha = 1/2) +
  xlab('Year') +
  ylab('Estimated Country BMI') +
  ggtitle('Evolution of Gender BMI in World Populations\n')

```

This scatterplot provides an interesting reflection on the evolution of the BMI
worldwide.

As a reminder, the current method for determining obesity classifies an 
individual as underweight if its BMI is below 18.5. Healthy individuals are
expected to have a BMI between 18.5 to 25, while overweight individuals comprise
BMIs between 25 and 30. Finally, an individual with a BMI greater than 30 is
classified as obese. 

Although some populations could be classified as underweight at the early 80s, 
no population average currently at this level. However, what is worrying is that
the 2008 BMI estimate for more than half of the countries populations was 
greater than 25: at an overweight level. Even more, some Pacific countries
population have BMI estimates greater than 30, which directly means that the
majority of the population on these countries is obese.

The health implications of this obesity epidemic must be (as they are) a medical
priority.

## Cholesterol blood levels and BMI are not correlated in worldwide populations.

```{r fig.width = 10, Final_Plot_2}

ggplot(aes(x = BMIClassification, y = Cholesterol), 
       data = nonPoor(healthData2008)) +
  geom_boxplot(aes(fill = Gender), alpha = 1/5) +
  xlab('Country BMI Classification') +
  ylab('Average country blood cholesterol levels (mmol/L)') +
  ggtitle('Relationship between BMI classification and Cholesterol blood levels
          in "non-poor" world populations. 2008\n') +
  geom_hline(yintercept=(5.2), linetype = 2, color = '#333333', alpha = 1/2)

```

I then wanted to study if I could correlate this BMI increase to a bad diet.
Although this information is difficult to address, some indirect indicators of 
diet type, such as cholesterol levels, may be interesting to analyze. Fast food
and bad dieting increase Cholesterol blood levels, so I thought that maybe there
was a correlation between the cholesterol levels and the BMI. 

One important consideration when analyzing this country data was the development
level of every country, because of the nutritional diferences between 
"westernized" and "non-westernized" countries (see 2.5.1). To take this into
account, I ignored the data from the bottom 25% "poorest" countries as
determined by its GDP (PPP). 

However, I did not find any meaningful relationship between the BMI and blood
Cholesterol levels ata population level.

## High Cholesterol blood levels are correlated with higher cancer cases.

```{r fig.width = 10, Final_Plot_3}

ggplot(aes(x = CholesterolRounded, y = meanNumberOfCancers, 
           fill = TypeOfCancer), data = AllCholCancerSummary) +
  geom_bar(stat = 'identity') +
  xlab('Population Cholesterol blood levels (mmol/L)') +
  ylab('Cancer cases per 100,000\nhabitants Male and Female Average') +
    ggtitle('Relationship between Cholesterol blood levels and Cancer cases 
            in mixed gender world populations. 2002\n') +
  geom_vline(xintercept = 5.2, linetype = 2, alpha = 1/3)

```

I finally analyzed the relationship between the different population BMI and
cholesterol levels and the incidence of different types of cancer. Although I 
did not found any special relationship between BMI and cancer incidence as I was
expecting, I found this relationship to be with the blood cholesterol levels. 

Higher cholesterol blood levels correlated with higher cancer incidences of
different types in both male and female populations (see 3.6) . As cholesterol 
in blood levels are highly related to the diet, my initial guess was that I 
would find more meaningful correlations between those levels and 
gastrointestinal cancers, but only found it to be true with Colon Cancer.

# 5. Reflection

The datasets used in this study contain information of different parameters from
about 200 world countries. Sometimes I struggled to maintain the reason because 
the data for male and female populations was separated. This enabled finer
exploration of the parameters, but also complicated a lot of otherwise simple
processes. I started exploring the principal parameter that I wanted to study: 
the population BMI, and could create a linear model for its worlwide increase. 

Eventually I integrated this data with the only diet-related parameter I could 
find in the Gapminder database: the total cholesterol blood levels. As I was
analyzing this relationship, I realized that I shouldhave into account the
differences between country development, and I also integrated data for each
country GDP (PPP). Unfortunately, I did not find the relationship I was 
expecting between BMI and Cholesterol levels, and I could not hint to a certain
effect of the bad dieting to this overweight and obesity increase worldwide.

Nevertheless, I continued my exploration to the main question: are those changes
in obesity affecting the cancer incidence? I decided to analyze the cholesterol
data too because it already was available, and found in this variable that the
correlation that I was searching in the BMI but didn't found. This relationship 
was specially highlighted in the Colon Cancer, where I draw a linear model of 
the correlation. I struggled in the normalization of the data, as I had to 
account for the number of cases each observation was made in order to depict a
correct ratio of cancer incidences to each parameter.

Finally, I'd like to talk a little about the data and conclusions that can be 
drawn from this study. I was interested in analyzing data regarding metabolism 
and health because studying the realtionship between the metabolic state and
inflammation was an important part of my PhD. The data used here is at the
epimidiologic level because working with individual data is protected and 
requires special permissions. However, to draw strong conclusions between the
factors studied here, the individual data is more suited. Moreover, we do not 
have available data for all types of cancer, and the time points are not 
complete. Not only that, but the analysis of one only timepoint is probably not
correct, as the timescale for the cancer development as a result of the 
different variables here studied should be taken into account. Finally, BMI is 
not perfect and more specific parameters may be better suited for this type of
study.